---
title: "Airline Punctuality & Delay Causes US"
author:
  - "Emil Sitka"
  - "Luise Killich"
  - "Ziqi Ang"
  - "Franka Konold"

format:
  dashboard:
    scrolling: true
    orientation: rows
    theme: vapor
date: today
---

# Introduction

 Introduction

::: {.callout-tip appearance="simple"}

# ✈️ Flight Delays in the United States

This dashboard explores monthly flight delays across the United States using publicly available airline performance data. Interactive visualisations and statistical analyses are used to investigate delay patterns across airlines, airports, and time.

:::

## 📊 Dataset

The dataset contains monthly summaries of airline performance collected by the U.S. federal government. Each observation represents the performance of a specific airline at a specific airport during a given month.

The data includes:

- Airline and airport identifiers
- Year and month
- Number of arriving flights
- Delayed, cancelled and diverted flights
- Delay durations by cause
- On-time performance measures
- Overall delay rates

> **Use the navigation tabs above to explore the analyses and visualisations.**


# RQ1: Patterns {.scrolling}
This Page explores descriptive statistics

::: {.panel-tabset}

## Spatial and Operational Patterns 



```{r}
#| title: "rq1_1 "



source("./code/01_download.R")
source("./code/02_clean.R")
source("./code/03_eda.R")
source("./code/04_analysis.R")

```


### Row {height=500px}

#### Column {width=50%}
```{r}
#| title: "Distribution of Delay Ratios by Cause"

p_rq1time_plot3
```

#### Column {width=50%}
```{r}
#| title: "Distribution of Carrier Delay Ratio"

p_rq1time_plot2
```

### Row {.tabset} {height=500px}

```{r}
#| title: "Q-Q Plot for Arrival Delays"


p_rq1_plot3
```

```{r}
#| title: "Distribution of all Arrival Delays"

p_rq1_plot2

```

### Row {height=500px}

#### Column {width=50%}
```{r}
#| title: "Delay Distribution of 10 biggest Airports compared to all Airports"

p_top_airports
```

#### Column {width=50%}
```{r}
#| title: "Delays by Carrier"

p_rq1_plot4
```


### Row {height=500px}

#### Column {width=50%}

```{r}
#| title: "Bottom 10 States by mean Arrival Delay"
p_bot10_states
```

#### Column {width=50%}
```{r}
#| title: "Top 10 States by mean Arrival Delay"
p_top10_states
```


### Row {height=500px}

::: {style="height:500px; overflow:hidden"}

```{r}
#| title: "Delay Distribution of all States"
#| fig-width: 12
#| fig-height: 5
#| out-width: "100%"
#| out-height: "100%"

p_state_boxplot
```

:::


## Temporal Patterns and Observations {.scrolling fill=false}



### Row {height=500px}

#### Column {width=50%}
```{r}
#| title: "Yearly Trend of all Delay Causes"

p_rq1time_plot1
```

#### Column {width=50%}
```{r}
#| title: "Detailed Yearly Trend"

fig
```

### Row {height=700px}

#### Column{width=50%}
```{r}
#| title: "Probability of Delays across Months"

p_rq1time_plot4
```

#### Column {width=50%}
```{r}
#| title: "Operational Delay Counts vs Total Arrival Delays: Quantile(τ=0.1) vs Mean Regressions"

p_rq1time_plot5
```

### Discussion

Observation:

With those Plots we can observe that the mean arrival delay is fluctuating in different years and also shows that some States or airports to a lesser extent also vary in the mean delay of flights arriving there but also in their variance. For example in the temporal development a particular drop can be seen during 2020 which can be attributed to the Corona Virus Pandemic. Also while in the case of airports there seems to be small differences, mainly in the deviation of the delays, when looking at States we can observe bigger differences where the mean Delay of flights landing in New jersey is more than double the mean delay in Montana or Idaho among others. 


:::

# RQ2: Correlations

## Row {height=500px}

### Column {width=50%}

```{r}
#| title: "Relation between total number of flights and number of delayed flights"

p_rq2_plot1

```

### Column{width=50%}

```{r}
#| title: "Relation between delay rate and total number of arriving flights"

p_rq2_plot2

```

## Observation 

Conclusion for RQ2:

Flight volume has a strong positive relationship with the absolute number
of delayed flights. However, it does not significantly affect the delay rate,
indicating that larger flight volumes lead to more delays in total but not
necessarily to a higher probability of delay.



# RQ3: Prediction {.scrolling}

## Row {height=500px}
```{r}
#| title: "Correlation Matrix of number of total flights,flights with delays and delay rate"
corrplot(cor_matrix,
        method = "color",
         addCoef.col = "black")

```

## Row {height=500px}

### Column{width=50%}
```{r}
#| title: "Regression Diagnostics for absolute Model"
par(mfrow = c(2, 2))
plot(model_absolut)

```

### Column{width=50%}
```{r}
#| title: "Regression Diagnostics for relative Model"
### Diagnostics for relative delay model
par(mfrow = c(2, 2))
plot(model_relative)

par(mfrow = c(1,1))
```

## Row {height=500px}

### Column {width=50%}

```{r}
#| title: "Predicted vs. Observed Values"

print(p_rq3_plot1)


```

### Column {width=50%}

```{r}
#| title: "Estimated Regression Coefficients"

p_rq3_plot2



```

## Results

<!-- Summary of key findings. -->

#### Results of RQ3:

Results:

The response variable showed substantial overdispersion (variance \> mean), making a quasi-Poisson regression more appropriate than a standard Poisson model.

Security delay had the strongest positive association with the monthly number of delayed flights, followed by the number of arriving flights. Weather delay, NAS delay, and late aircraft delay showed relatively small positive effects, while carrier delay had a slight negative coefficient after controlling for the other predictors.

VIF values ranged from 2.27 to 5.80, indicating low to moderate multicollinearity without evidence of severe collinearity.

The observed-versus-predicted plot shows that the model captures the overall trend but produces larger prediction errors for observations with very high delay counts.

## Discussion

<!-- Interpret results, limitations, and next steps. -->

#### Discussion for RQ3: 

Discussion:

The regression results indicate that operational delay measures are associated with the number of delayed flights, but these relationships should not be interpreted as causal effects because both predictors and the response were derived from the same monthly operational data.

The negative coefficient for carrier delay likely reflects overlap and shared variation with the other delay variables rather than a true negative effect.

Although multicollinearity was present, the VIF values suggest that it is unlikely to substantially affect the overall model. The model explains the general pattern of delayed flights but is less accurate for extreme observations, indicating that additional factors influence unusually high delay counts.
