---
title: "Project Proposal"
author:
  - "Emil Sitka"
  - "Luise Killich"
  - "Ziqi Ang"
  - "Franka Konold"
date: today
editor: 
  markdown: 
    wrap: 72
---

## BTS \| OT Delay

<!-- Describe the dataset: source, licence, dimensions, key variables. -->

### Source

The dataset used in this project is provided by the Bureau of
Transportation Statistics (BTS), part of the U.S. Department of
Transportation (USDOT). Specifically, it originates from the Airline
Service Quality Performance (ASQP) database. Under federal regulations
air carriers with at least 1% of domestic scheduled-passenger revenues
are legally required to report monthly on-time performance and detailed
flight delay causes to the BTS.

### Licence

As a U.S. federal government dataset, it is released into the public
domain (U.S government Data). This framework allows us to download,
analyse, modify, and publish results based on the dataset without
requiring a license or proprietary permissions.

### Structure of the dataset

The dataset is structured as a monthly summary grouped by airline and
airport. Each row in the dataset gives information about how a specific
carrier performed at a specific airport during a specific month and
year. It contains data from 2003 until 2026 for 450 of the 503 airports
that serve commercial flights in the US.

The 25 variables can be grouped into four categories:

- Identifiers: Who, where and when (`year`, `month`, `carrier`,
  `carrier_name`, `airport`, `airport_name`)

- Headcounts: the raw number of flights that scheduled to land, how many
  were on time and how many ended up cancelled or diverted
  (`arr_flights`, `arr_del15`, `arr_cancelled`, `arr_diverted`,
  `ontime_flights`)

- Responsibility: The number of flights assigned to a specific delay
  category (`carrier_ct`, `weather_ct`, `nas_ct`, `security_ct`,
  `late_aircraft_ct`)

- Rates: The actual duration of the delays in minutes and the calculated
  rates to show the percentage of flights on time vs late (`delay_rate`,
  `onetime_rate` and `avg_delay_per_delayed_flight`)

Additionally, when a single flight experiences multiple sources of
delay, the dataset allocates both the minutes and flight counts
proportionally. This ensures that the sum of all fractional count
variables within any given row equals the total integer count of delayed
flights.

For example, if a flight is delayed 60 minutes, with 30 minutes due to
carrier issues and 30 minutes due to weather, both `carrier_delay` and
`weather_delay` receive 30 minutes, while `carrier_ct` and `weather_ct`
each receive a fractional count of 0.5.

### Data Dictionary

| Variable Name | Data Type | Description |
|:-----------------|:-----------------|:------------------------------------|
| **year** | double | Year of the flight. |
| **month** | double | Month of the flight. |
| **carrier** | char | Unique two-letter airline carrier code |
| **carrier_name** | char | The full commercial name of the airline carrier. |
| **airport** | char | The 3-letter IATA airport code (e.g. JFK, LAX). |
| **airport_name** | char | The full name of the airport along with its city and state location. |
| **arr_flights** | double | Total number of arriving flights for that carrier/airport/month combination. |
| **arr_del15** | double | Total number of arriving flights delayed by 15 minutes or more. |
| **carrier_ct** | double | The fractional count of flights delayed due to airline-controlled factors. |
| **weather_ct** | double | The fractional count of flights delayed due to significant or extreme weather conditions. |
| **nas_ct** | double | The fractional count of flights delayed due to the National Airspace System. |
| **security_ct** | double | The fractional count of flights delayed due to airport security breaches or issues. |
| **late_aircraft_ct** | double | The fractional count of flights delayed because the incoming aircraft arrived late. |
| **arr_cancelled** | double | Total number of arriving flights that were cancelled. |
| **arr_diverted** | double | Total number of arriving flights that were diverted to an alternate airport. |
| **arr_delay** | double | Total arrival delay minutes accumulated by all flights. |
| **carrier_delay** | double | The cause of delay was within airline´s control (e.g. maintenance, fueling, etc.) |
| **weather_delay** | double | Meteorological conditions (actual or forecasted) that prevents operation of a flight. |
| **nas_delay** | double | Delays attributable to the national aviation system that refer to conditions as non-extreme weather conditions, airport operations, heavy traffic volume and air traffic control. |
| **security_delay** | double | Delays caused by evacuation of a terminal or concourse, reboarding of aircraft because of security breach, inoperative screening equipment and/or long lines in excess of 29 minutes at screeing areas. |
| **late_aircraft_delay** | double | A previous flight with same aircraft arrived late, causing the present flight to depart late. |
| **ontime_flights** | double | Flights are considered ontime if it operates less than 15 minutes later than the scheduled time. |
| **delay_rate** | double | The proportion of delayed flights relative to total arriving flights (`arr_del15` / `arr_flights`). |
| **ontime_rate** | double | The proportion of on-time flights relative to total arriving flights (`ontime_flights` / `arr_flights`). |
| **avg_delay_per_delayed_flight** | double | The average duration of delays in minutes, calculated only for flights that were actually delayed (`arr_delay`/ `arr_del15`). |

## Research Questions

1.  How do flight delays vary across airlines, airports and over time?
    What is the distribution of different delay causes across flights,
    airlines and airports? What temporal and seasonal patterns can be
    observed?

2.  Is there a relationship between the number of flights (arr_flights)
    and the number of delayed flights (arr_del15)? Can we predict total
    arrival delays based on flight volume?

3.  Can arrival delays (arr_del15) be predicted based on flight
    characteristics and different delay causes such as carrier, weather,
    NAS, security, and late aircraft delays? What factors are
    significant predictors of flight delays?

## Context and Audience

This analysis is conducted in an academic setting as part of the
Advanced Statistical Programming course.

The primary target audience is lecturers.

The results are presented as an interactive quarto report that combines
code, visualisations and text for data storytelling and reproducible
analysis.

## Initial EDA

### Data overview

```{r}
#| label: load-data
library(tidyverse)
library(visdat)
library(skimr)
library(here)
source("./code/02_clean.R")

glimpse(delay)

vis_dat(head(delay, 500))

skim(select(delay,
            arr_flights,
            arr_del15,
            arr_delay))
```

### Data quality checks

Due to the large size of the dataset, a random sample of 55,000
observations was taken.

- Variables are correctly typed overall, with month potentially better
  treated as a categorical variable.

- In one row (40418), negative values were identified in `nas_ct`(-0.01)
  and `nas_delay`(-1). Both values were treated as missing values (NA).

- Delay-cause variables such as `carrier_ct` contain decimal values
  because they represent allocated delay causes rather than strict
  counts.

- Several delay-related variables contain missing values (664–964
  observations), which appear to be structurally driven.

- No obvious inconsistencies were found in categorical variables.

### Data types

This dataset contains:

- integer variables: `year`, `month`, `arr_flights`, `arr_del15`,
  `arr_cancelled`, `arr_diverted`, `arr_delay`, `carrier_delay`,
  `weather_delay`, `nas_delay`, `security_delay`, `late_aircraft_delay`

- numeric variables: `carrier_ct`, `weather_ct`, `nas_ct`,
  `security_ct`, `late_aircraft_ct` (do we need to change them to
  double?)

- character variables: `carrier`, `carrier_name`, `airport`,
  `airport_name`

### Count NA entries, diagnose missingness mechanism MAR, MCAR, MNAR, decide solution

### Visual Exploration

```{r}
library(ggplot2)

# Plot Distribution of arrival delays
ggplot(delay, aes(x = arr_delay)) +
  geom_histogram(bins = 50) +
  labs(title = "Distribution of arrival delays",
       x = "Arrival delay (minutes)", y = "Count")

```

The distribution of arrival delays is strongly right-skewed. Most
flights experience small delays, while a small number of extreme delays
create a long right tail.

```{r}
# Plot relationship between flight volume and delays
ggplot(delay, aes(x = arr_flights, y = arr_del15)) +
  geom_point(alpha = 0.6) +
  labs(title = "Relationship between number of flights and delayed flights",
       x = "Number of flights",
       y = "Number of delayed flights")
```

The scatterplot suggests a positive relationship between flight volume
and delayed flights, indicating that higher traffic is associated with
more delays.

```{r}
library(dplyr)
library(tidyverse)
delay <- delay %>%
  mutate(
    ontime_flights = arr_flights - arr_del15,
    delay_rate = arr_del15 / arr_flights,
    ontime_rate = ontime_flights / arr_flights,
    avg_delay_per_delayed_flight = ifelse(
      arr_del15 > 0,
      arr_delay / arr_del15,
      NA
    )
  )
summary(delay$delay_rate)
mean(delay$delay_rate, na.rm = TRUE)

# compute year delay rate over years
yearly <- delay %>%
  group_by(year) %>%
  summarise(
    total_flights = sum(arr_flights, na.rm = TRUE),
    total_delayed = sum(arr_del15, na.rm = TRUE),
    delay_rate = total_delayed / total_flights
  )

# plot the overall trend
ggplot(yearly, aes(x = year, y = delay_rate)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = yearly$year[seq(1, length(yearly$year), by = 2)]) +
  labs(
    title = "Flight Delay Rate Over Time",
    x = "Year",
    y = "Delay Rate"
  )
```

Trend analysis, delay rate over year, (dataset is not unified with other
members, because the dataset needs to be discussed with other members
about how to split under 10mb)

<!-- Add 2–3 exploratory plots and a brief commentary on what you observe. -->

## Summary Statistics

- Central Tendency: calculate the mean, median, and mode
- Spread and Variability: standard deviation, variance, and range
- test data normality with QQ plot
- data distribution with Histogram

## Analysis Plan

RQ1 - descriptive statistics and group comparison (Boxplots, means) -
compare `arr_delay` across `carrier` - plots over
airlines/airports/time - trend analysis over year, detect anomalies
(e.g.sudden drop in 2020 & Covid Pandemie) - cluster delay type, plots
line charts

RQ2 - scatterplots - visualize relationship between two variable to
identify correlation - Heatmaps - correlation among multiple variables -
correlation analysis

RQ3 - Regression - predictors: flights + delay causes - dataset
observation: `arr_delay`= `carrier_delay`+ `weather_delay`+
`nas_delay`+`security_delay`+`late_aircra`; `arr_del15` als Zähldaten
(can we do linear regression or possion regression?)

<!-- Outline the methods you plan to use to answer your research questions. -->
